Lazy Linearity for a Core Functional Language

### 2025-10-20 popl26.tex
revised and fit to 25'

Linear Mini-Core~\cite{cite:minicore} is a specification of linear types in
Core as they were being implemented in GHC, and doubles as the (unpublished)
precursor to our work. Linear Mini-Core first observes the incapacity of
Core's type system to accept linear programs after transformations, and 
introduces usage environments for let-bound variables with the same goal of
Linear Core of specifying a linear type system for Core that accepts the
optimising transformations.
%
We draw from Linear Mini-Core the rule for non-recursive let expressions and
how let-bound variables are annotated with a usage environment. However, our
work further explores the interaction of laziness with linearity in depth, and
diverges significantly in rules for typing other constructs, notably, case expressions and
case alternatives. Furthermore, unlike Mini-Core, we prove Linear Core is type
safe, guarantees linear resource usage, and that multiple optimising
transformations, when applied to Linear Core programs, preserve linearity.
\ROUNDTWO{Our irrelevant resources
  have similarities with 0-multiplicity
  values from QTT~\cite{brady:LIPIcs.ECOOP.2021.9} and, as previously noted,
  our tagged resources are related to fractional permissions from separation
  logic~\cite{10.5555/1760267.1760273}.}

@@

\paragraph{Linearity-directed optimisations}

Core-to-Core transformations appear in multiple papers~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
all designed in the context of a typed language (Core) which does not have
linear types. However, some authors~\cite{cite:let-floating,peytonjones1997a,10.1145/3158093} have observed that
optimisations (in particular, let-floating and inlining) can greatly
benefit from linearity analysis and, in order to improve those transformations,
special purpose linear-type-inspired systems were created and implemented.
%
Preserving linear types in Core throughout the compilation pipeline allows the
optimiser to leverage non-heuristic linearity information where, previously, it
would resort solely to ad-hoc or incomplete custom-built linearity inference
passes (naturally, these passes are still necessary to optimise programs not
using explicit linearity or multiplicity annotations). \ROUNDTWO{Linearity may
potentially be used to the benefit of other transformations.
An obvious candidate is
\emph{inlining}, which is applied based on heuristics from information provided
by the \emph{cardinality analysis} pass that counts occurrences of bound
variables.  Linearity can be used to non-heuristically inform
the inliner~\cite{10.1145/3158093}.}

@@

We highlight some avenues of future work. Briefly,
these include \emph{multiplicity coercions}, optimisations leveraging
linearity, resource inference for usage environments, and ultimately using
Linear Core in a mature optimising compiler with lazy evaluation and linear
types -- the Glasgow Haskell Compiler. Lastly, we discuss the
% we hypothesize Linear($X$), a linear system parametrized on the evaluation strategy, and the
generalization of Linear Core to the surface Haskell language.

>>>>

Linear Mini-Core~\cite{cite:minicore} is a specification of linear types in
Core as they were being implemented in GHC, and doubles as the (unpublished)
precursor to our work. Linear Mini-Core first observes the incapacity of
Core's type system to accept linear programs after transformations, and 
introduces usage environments for let-bound variables with the same goal of
Linear Core of specifying a linear type system for Core that accepts the
optimising transformations.
%
We draw from Linear Mini-Core the rule for non-recursive let expressions and
how let-bound variables are annotated with a usage environment. However, our
work further explores the interaction of laziness with linearity in depth, and
diverges significantly in rules for typing other constructs, notably, case expressions and
case alternatives. Furthermore, unlike Mini-Core, we prove Linear Core is type
safe, guarantees linear resource usage, and that multiple optimising
transformations, when applied to Linear Core programs, preserve linearity.
\ROUNDTWO{Our irrelevant resources
  have similarities with 0-multiplicity
  values from QTT~\cite{brady:LIPIcs.ECOOP.2021.9} which are
  disallowed from being used directly in executable code and, as previously noted,
  our tagged resources are related to fractional permissions from separation
  logic~\cite{10.5555/1760267.1760273}.}

@@

\paragraph{Linearity-directed optimisations}
%
Core-to-Core transformations appear in multiple papers~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
all designed in the context of a typed language (Core) which does not have
linear types. However, some authors~\cite{cite:let-floating,peytonjones1997a,10.1145/3158093} have observed that
optimisations (in particular, let-floating and inlining) can greatly
benefit from linearity analysis and, in order to improve those transformations,
special purpose linear-type-inspired systems were created and implemented.
%
Preserving linear types in Core throughout the compilation pipeline allows the
optimiser to leverage non-heuristic linearity information where, previously, it
would resort solely to ad-hoc or incomplete custom-built linearity inference
passes (naturally, these passes are still necessary to optimise programs not
using explicit linearity or multiplicity annotations). \ROUNDTWO{Linearity may
potentially be used to the benefit of other transformations.
An obvious candidate is
\emph{inlining}, which is applied based on heuristics from information provided
by the \emph{cardinality analysis} pass that counts occurrences of bound
variables.  Linearity can be used to non-heuristically inform
the inliner~\cite{10.1145/3158093}.}

@@

We highlight some avenues of future work. Briefly,
these include \emph{multiplicity coercions}, optimisations leveraging
linearity, %resource inference for usage environments,
and incorporating Linear Core in GHC Haskell. % Lastly, we discuss the
% % we hypothesize Linear($X$), a linear system parametrized on the evaluation strategy, and the
% generalization of Linear Core to the surface Haskell language.
